Improving Item-Item Similarity Estimation during Collaborative Filtering

### 2014-03-25 paper/paper.tex
pulling
Extend model-based section.
Develop model-based section.
Changed names of sections. Edited probabilstic model section of ratings model

\newcommand{\spc}{\vspace{0.25in}}      % Shortcut commands
\newcommand{\ds}{\displaystyle}         %\newcommand{\ds}[1]{\displaystyle{#1}}
\newcommand{\ra}{\rightarrow}
\DeclareMathOperator*{\argmax}{arg\!\max}
\DeclareMathOperator*{\argmin}{arg\!\min}

@@

\section*{True Item-Item Similarity}

@@

\section*{Estimating Item-Item Similarity}

@@

\section*{Linear Approach}
\subsection*{Probabilistic Model}

@@

We model $Y$ as a Normal distribution 
\begin{align}
Y \sim N(\mu, \sigma_{1}^2)
\end{align}
where $\mu$ and $\sigma_{1}^2$ are the average similarity score and the variance 
of similarity scores of all pairs of items, respectively. Since $X_n$ represents
a noisy reading of the true similarity $Y$, we model $X_n | Y=y$ as a Gaussian
error around $y$:
\begin{align}
(X_n | Y=y) \sim N(y, \sigma_{2, n}^2)
\end{align}

Note that $\sigma_{2, n}^2$ represents how noisy $Sim_n$ is as an estimation of
$TrueSim$ and therefore should decrease as $n$ increases. The probability
distribution of $X_n$ can be visualized in Figure 1.

@@

The problem of approximating the $TrueSim$ from $Sim_n$ can now be seen as
finding the value of $Y$ that most likely produced $X_n$. In other words we 
desire the Maximum Likelihood Estimate of $Y | X_n$:
\begin{align}
\hat{y} &= \argmax_yP(Y=y|X_n=x) 
\\&= \argmax_y\left[\frac{P(X_n=x | Y=y)P(Y=y)}{P(X_n=x)}\right]
\\&= \argmax_y\left[P(X_n=x|Y=y)P(Y=y)\right]
\\&= 
\argmax_y\left[\frac{1}{\sigma_{2,n}\sqrt{2\pi}}\exp{\left(\frac{-(x-y)^2}
{2\sigma_{2,n}^2}\right)}
\frac{1}{\sigma_{1}\sqrt{2\pi}}\exp{\left(\frac{-(y-\mu)^2}
{2\sigma_{1}^2}\right)}\right]
\\&= \argmin_y\left[\frac{(x-y)^2}{2\sigma_{2,n}^2} +
\frac{(y-\mu)^2}{2\sigma_{1}^2}\right]
\\&= \argmin_y\left[\left(\sigma_{1}^2+\sigma_{2,n}^2\right)y^2 - 
2\left(\sigma_{1}^2x+\sigma_{2,n}^2\mu\right)y\right]
\end{align}
We find the exact minimum by taking the derivative with respect to $y$ and
setting it to zero:
\begin{align}
\frac{d}{dy}\left[\left(\sigma_{1}^2+\sigma_{2,n}^2\right)y^2 - 
2\left(\sigma_{1}^2x+\sigma_{2,n}^2\mu\right)y\right]
&= 2\left(\sigma_{1}^2+\sigma_{2,n}^2\right)y - 
2\left(\sigma_{1}^2x+\sigma_{2,n}^2\mu\right) 
\\&= 0
\end{align}
This produces the following linear equation for \hat{y} in terms of x:
\begin{align}
\hat{y} &= \frac{\sigma_{1}^2x+\sigma_{2,n}^2\mu}{\sigma_{1}^2+\sigma_{2,n}^2}
\end{align}
which can be rewritten as:
\begin{align}
\left(\hat{y} - \mu\right) &= \frac{\sigma_{1}^2}{\sigma_{1}^2+\sigma_{2,n}^2}
\left(x-\mu\right)
\end{align}

The model suggests that there is a linear correlation between $X_n$ ($Sim_n$)
and $Y$ ($TrueSim$). Moreover, the best predicted $TrueSim$ is a linear 
combination of the mean and the observed similiarity. Since $\sigma_{2,n}^2$ is 
nonnegative, the slope is always less than or equal to one, meaning that the 
observed distance from $\mu$ (the average true similarity) overestimates the 
true distance. This is reasonable, since there is a prior that expects the true 
similarity to be around $\mu$.

The parameters $\mu$, $\sigma_{1}^2$, and $\sigma_{2,n}^2$ can be approximated 
using the method of moments over training data where there are enough common
users that $TrueSim$ can be approximated. This method requires approximating
$\sigma_{2,n}^2$ for every value of $n$ separately. To gain a more general 
model, we can model $\sigma_{2,n}^2$ as a function of $n$. Recall that 
$\sigma_{2,n}^2$ is the noisiness of the measurement $Sim$ about $TrueSim$ and $TrueSim = lim_{n \to
\infty}Sim$. Thus $Sim$ can be thought of as a sampling $n$ users from the
infinite set used to calculate $TrueSim$. Although $Sim$ may not be linear, we
the Central Limit Theorem motivates the intuition that the variance of $Sim$
would decrease as the $sqrt(n)$. Thus we model $\sigma_{2,n}^2$ as:
\begin{align}
\sigma_{2,n}^2 = \frac{\alpha}{\sqrt{n}},
\end{align}
yielding the single parameter model:
\begin{align}
\left(Y - \mu\right) = \frac{\sigma_{1}^2}{\sigma_{1}^2+\frac{\alpha}{\sqrt{n}}}
\left(X_n-\mu\right)
\end{align}
In terms of a modified similarity function, as desired earlier, we now have:
\begin{align}
ModSim(Sim, n) = \frac{\sigma_{1}^2}{\sigma_{1}^2+\frac{\alpha}{\sqrt{n}}}
\left(Sim-\mu\right) + \mu
\end{align}

\section*{Technique}
Based on this maodel, we try three different techniques to predict the true similarity from a
measured similarity. The techniques vary in their fidelity to the model. For
each, we limit our data to all pairs of items that have enough common users that
we can approximate $TrueSim \approx Sim$. Then we pick subsets of the common
users of various sizes $n$ and calculate $Sim_{n}$. With examples of $Sim$, $n$
and the resulting $TrueSim$, we can implement supervised learning.

The first technique is the least faithful to the model. It simply runs a number
of linear regressions between $Sim$ and $TrueSim$ for each $n$. This ignores the 
data-specific parameters $\mu$, $\sigma_{1}$ and $\sigma_{2,n}$ to find the best 
linear fit. It also treats each $n$ completely independently. This technique 
necessarilty has the lowest training error of all linear techniques, potentially 
at the risk of overfitting.

The second technique is more general. It approximates $\mu$ and $\sigma_{1}$
using the method of moments on the training data of all $TrueSim$ scores. Then,
$\sigma_{2,n}$ is also approximated using a method of moments on the training
data of all $Sim_n$ for each $n$ separately. Here $\sigma_{2,n}$'s are also
considered independently of each other.

The third and most general technique models $\sigma_{2,n}$ as a function of
$n$ using equation 19. As in the previous technique, $\mu$, $\sigma_{1}$ and each
$\sigma_{2,n}$ are approximated using the method of moments over the training
data. Then $\alpha$ is calculated that minimizes the sum of squared errors
between $\sigma_{2,n}^2$ and $\frac{\alpha}{\sqrt{n}}$. This $\alpha$ is used to
construct the three-parameter prediction function of equation 19.

\section*{A Model-based Technique}

We now construct a relatively simple probabilistic model that is readily
exploited for making estimates of $TrueSimRand$ and, to a lesser extent,
$TrueSimPref$. In addition, this approach has the added benefit of allowing us
to infer implied predictions of $TrueSimRand$ made by individual users each time
they rate two items. This latter feature allows us to define a new metric for
users that we call {\em user-predictivity}, which allows us to ascribe varying
levels of confidence to their contributions to a similarity estimate.

Let $Y$ be a random variable representing the $TrueSimRand$ of a randomly chosen
pair of items. As before, we model $Y$ as a Normal distribution
\begin{align}
Y \sim N(\mu_s, \sigma_s^2),
\end{align}
where $\mu_s$ and $\sigma_s$ are the average and standard deviation of all
similarity scores, respectively. Now, let $\vec{R} = \left(R_A~~R_B\right)$ be
a random two element vector representing the ratings given to items $A$ and $B$.
We model the conditional distribution of $\vec{R} | Y$ as a multivariate Normal
distribution
\begin{align}
\vec{R} | Y=y \sim N(\vec{\mu_r}, \mathbf{\Sigma_y}),
\end{align}
where the elements of $\vec{\mu_r}$ are the average ratings and
$\mathbf{\Sigma_y}$ is the $y$-dependent $2\times2$ covariance matrix.

>>>>

\newcommand{\spc}{\vspace{0.25in}}      % Shortcut commands
\newcommand{\ds}{\displaystyle}         %\newcommand{\ds}[1]{\displaystyle{#1}}
\newcommand{\ra}{\rightarrow}
\newcommand{\Cov}{\mathrm{Cov}}
\DeclareMathOperator*{\argmax}{arg\!\max}
\DeclareMathOperator*{\argmin}{arg\!\min}

@@

\section*{Similarity}
\subsection*{True Item-Item Similarity}

@@

\subsection*{Estimating Item-Item Similarity}

@@

\section*{Score-based Approach}
\subsection*{Probabilistic Model}

@@

We model $Y$ as a Normal distribution 
\begin{align}
Y \sim N(\mu, \sigma_{s}^2)
\end{align}
where $\mu$ and $\sigma_{s}^2$ are the average similarity score and the variance 
of similarity scores of all pairs of items, respectively. Since $X_n$ represents
a noisy reading of the true similarity $Y$, we model $X_n | Y=y$ as a Gaussian
error around $y$:
\begin{align}
(X_n | Y=y) \sim N(y, \sigma_{err, n}^2)
\end{align}

Note that $\sigma_{err, n}^2$ represents how noisy $Sim_n$ is as an estimation of
$TrueSim$ and therefore should decrease as $n$ increases. The probability
distribution of $X_n$ can be visualized in Figure 1.

@@

The problem of approximating the $TrueSim$ from $Sim_n$ can now be seen as
finding the value of $Y$ that most likely produced $X_n$. In other words we 
desire the Maximum Likelihood Estimate of $Y | X_n$:
\begin{align}
\hat{y} &= \argmax_yP(Y=y|X_n=x) 
\\&= \argmax_y\left[\frac{P(X_n=x | Y=y)P(Y=y)}{P(X_n=x)}\right]
\\&= \argmax_y\left[P(X_n=x|Y=y)P(Y=y)\right]
\\&= 
\argmax_y\left[\frac{1}{\sigma_{err,n}\sqrt{2\pi}}\exp{\left(\frac{-(x-y)^2}
{2\sigma_{err,n}^2}\right)}
\frac{1}{\sigma_{s}\sqrt{2\pi}}\exp{\left(\frac{-(y-\mu)^2}
{2\sigma_{s}^2}\right)}\right]
\\&= \argmin_y\left[\frac{(x-y)^2}{2\sigma_{err,n}^2} +
\frac{(y-\mu)^2}{2\sigma_{s}^2}\right]
\\&= \argmin_y\left[\left(\sigma_{s}^2+\sigma_{err,n}^2\right)y^2 - 
2\left(\sigma_{s}^2x+\sigma_{err,n}^2\mu\right)y\right]
\end{align}
We find the exact minimum by taking the derivative with respect to $y$ and
setting it to zero:
\begin{align}
\frac{d}{dy}\left[\left(\sigma_{s}^2+\sigma_{err,n}^2\right)y^2 - 
2\left(\sigma_{s}^2x+\sigma_{err,n}^2\mu\right)y\right]
&= 2\left(\sigma_{s}^2+\sigma_{err,n}^2\right)y - 
2\left(\sigma_{s}^2x+\sigma_{err,n}^2\mu\right) 
\\&= 0
\end{align}
This produces the following linear equation for $\hat{y}$ in terms of $x$:
\begin{align}
\hat{y} &= \frac{\sigma_{s}^2x+\sigma_{err,n}^2\mu}{\sigma_{s}^2+\sigma_{err,n}^2}
\end{align}
which can be rewritten as:
\begin{align}
\left(\hat{y} - \mu\right) &= \frac{\sigma_{s}^2}{\sigma_{s}^2+\sigma_{err,n}^2}
\left(x-\mu\right)
\end{align}

The model suggests that there is a linear correlation between $X_n$ ($Sim_n$)
and $Y$ ($TrueSim$). Moreover, the best predicted $TrueSim$ is a linear 
combination of the mean and the observed similiarity. Since $\sigma_{err,n}^2$ is 
nonnegative, the slope is always less than or equal to one, meaning that the 
observed distance from $\mu$ (the average true similarity) overestimates the 
true distance. This is reasonable, since there is a prior that expects the true 
similarity to be around $\mu$.

The parameters $\mu$, $\sigma_{s}^2$, and $\sigma_{err,n}^2$ can be approximated 
using the method of moments over training data where there are enough common
users that $TrueSim$ can be approximated. This method requires approximating
$\sigma_{err,n}^2$ for every value of $n$ separately. To gain a more general 
model, we can model $\sigma_{err,n}^2$ as a function of $n$. Recall that 
$\sigma_{err,n}^2$ is the noisiness of the measurement $Sim$ about $TrueSim$ and $TrueSim = lim_{n \to
\infty}Sim$. Thus $Sim$ can be thought of as a sampling $n$ users from the
infinite set used to calculate $TrueSim$. Although $Sim$ may not be linear, we
the Central Limit Theorem motivates the intuition that the variance of $Sim$
would decrease as the $sqrt(n)$. Thus we model $\sigma_{err,n}^2$ as:
\begin{align}
\sigma_{err,n}^2 = \frac{\alpha}{\sqrt{n}},
\end{align}
yielding the single parameter model:
\begin{align}
\left(Y - \mu\right) = \frac{\sigma_{s}^2}{\sigma_{s}^2+\frac{\alpha}{\sqrt{n}}}
\left(X_n-\mu\right)
\end{align}
In terms of a modified similarity function, as desired earlier, we now have:
\begin{align}
ModSim(Sim, n) = \frac{\sigma_{s}^2}{\sigma_{s}^2+\frac{\alpha}{\sqrt{n}}}
\left(Sim-\mu\right) + \mu
\end{align}

\subsection*{Technique}
Based on this maodel, we try three different techniques to predict the true similarity from a
measured similarity. The techniques vary in their fidelity to the model. For
each, we limit our data to all pairs of items that have enough common users that
we can approximate $TrueSim \approx Sim$. Then we pick subsets of the common
users of various sizes $n$ and calculate $Sim_{n}$. With examples of $Sim$, $n$
and the resulting $TrueSim$, we can implement supervised learning.

The first technique is the least faithful to the model. It simply runs a number
of linear regressions between $Sim$ and $TrueSim$ for each $n$. This ignores the 
data-specific parameters $\mu$, $\sigma_{s}$ and $\sigma_{err,n}$ to find the best 
linear fit. It also treats each $n$ completely independently. This technique 
necessarilty has the lowest training error of all linear techniques, potentially 
at the risk of overfitting.

The second technique is more general. It approximates $\mu$ and $\sigma_{s}$
using the method of moments on the training data of all $TrueSim$ scores. Then,
$\sigma_{err,n}$ is also approximated using a method of moments on the training
data of all $Sim_n$ for each $n$ separately. Here $\sigma_{err,n}$'s are also
considered independently of each other.

The third and most general technique models $\sigma_{err,n}$ as a function of
$n$ using equation 19. As in the previous technique, $\mu$, $\sigma_{s}$ and each
$\sigma_{err,n}$ are approximated using the method of moments over the training
data. Then $\alpha$ is calculated that minimizes the sum of squared errors
between $\sigma_{err,n}^2$ and $\frac{\alpha}{\sqrt{n}}$. This $\alpha$ is used to
construct the three-parameter prediction function of equation 19.

\section*{Ratings-based Approach}
\subsection*{Probabilistic Model}

We now construct a different probabilistic model based on the pairs of ratings
themselves. This model is readily exploited for making estimates of 
$TrueSimRand$ and, to a lesser extent, $TrueSimPref$. In addition, this approach 
has the added benefit of allowing us to infer implied predictions of 
$TrueSimRand$ made by individual users each time they rate two items. This 
latter feature allows us to define a new metric for users that we call {\em 
user-predictivity}, which allows us to ascribe varying levels of confidence to 
their contributions to a similarity estimate.

Let $Y$ be a random variable representing the $TrueSimRand$ of a randomly chosen
pair of items. As before, we model $Y$ as a Normal distribution
\begin{align}
Y \sim N(\mu_s, \sigma_s^2),
\end{align}
where $\mu_s$ and $\sigma_s^2$ are the average and variance of all true similarity
scores, respectively. Now, let $\vec{R} = \left(R_A~~R_B\right)$ be a random two
element vector representing the ratings given to items $A$ and $B$.  We model
the conditional distribution of $\vec{R} | Y$ as a multivariate Normal
distribution
\begin{align}
(\vec{R} | Y=y) \sim N(\vec{\mu_r}, \mathbf{\Sigma_y}),
\end{align}
where the elements of $\vec{\mu_r}$ are the average ratings and
$\mathbf{\Sigma_y}$ is the $y$-dependent $2\times2$ covariance matrix. 
Since $y$ is the $TrueSimRand$ score, it can be thought of as the Pearson 
correlation over the distribution of ratings $R_A | Y$ and $R_B | Y$. Assuming
that the variance of individual ratings $\sigma_r^2$ is independent of $y$, we have 
\begin{align}
y = \frac{\Cov{\left[R_A | Y=y , R_B | Y=y\right]}}{\sigma_r^2},
\end{align}
It therefore follows that if the ratings are indeed normally distributed 
then the covariance matrix is related to the true similarity $y$ by
\begin{align}
\mathbf{\Sigma_y} = \sigma_r^2 \cdot
\left[ \begin{array}{cc}  1 & y \\ y & 1 \end{array} \right]
\end{align}
